miniKanren with fair search strategies

### 2019-03-31 Paper/paper.tex
more intro

With interleaving depth-first search, disjunctive operator allocates half
computational effort to its first goal, and quater to the second, eighth to the
third, and so on. If a goal that can generate planty of unwanted partial 
solutions is placed at the beginning of a \texttt{conde} expressions, it might
waste a lot computational resource. Consequently, miniKanren programmers 
sometimes need to order their \texttt{conde} clauses carefully to prevent some 
clauses absorbing unproportional computational effort. For example, when a 
relational proof checker is used as a proof generator, the eliminational rules 
should comes later.
%TODO KC: prefer an example from other literatures
Seasoned miniKanreners usually know how to utilize the unfairness to optimize
their programs. However, we believe search strategies less sensitive to goal
order can also be useful to little miniKanreners as well as seansoned ones. We 
propose two such search strategies, balanced interleaving DFS (biDFS) and
breadth-first search (BFS), and observe how they affect the efficiency and
query result of known miniKanren programs. The experiment is conducted with the
miniKanren from \textit{The Reasoned Schemer, 2nd Edition}.

%TODO KC: Should we use microKanren instead?

%The version of miniKanren in \textit{The Reasoned Schemer, 2nd Edition} features
%an efficient and complete search strategy -- interleaving depth-first search (iDFS).
%iDFS biases toward left \texttt{conde} lines. So miniKanren programmers sometimes need to
%organize their \texttt{conde} lines carefully. We propose and implement two search strategies,
%breadth-first search (BFS) and balanced interleaving DFS (biDFS). With either of these
%search strategy, miniKanren becomes less sensitive to the order of \texttt{conde} lines.

(TODO: summarize following sections) Section 2 explains why iDFS biases toward left \texttt{conde} clauses. Section 3 is about the biDFS. Section 4 is about BFS.

\section{why iDFS put more effort computing left clauses}

\texttt{conde} clauses are combined by \texttt{disj2}, right associatively. The
core functionality of \texttt{disj2} is completed by \texttt{append-inf} (Fig.~
ref{disj2-and-append-inf}). When both input streams are infinite, the resulting
mature part contains only the mature part of \texttt{s-inf}. The whole
\texttt{t-inf} goes to the resulting immature part. However, \texttt{t-inf} and
\texttt{s-inf} are swapped in the delayed recursive call. Hence the search
strategy spend computational effort evenly in two appended streams. As
\texttt{disj2} is right associative, the left clauses are nested in less calls
to \texttt{append-inf}, which implies that they are more likeyly being
computed.

@@

In this section we change the search strategy to breadth-first search and 
optimize it. The whole process is completed in two steps. In the first step, 
from mk-0 to mk-1, BFS is introduced. In the second step, mk-1 to mk-3, BFS is 
optimized. The initial version, \texttt{mk-0}, is exactly the version in 
\emph{The Reasoned Schemer, 2nd Edition}.

@@

The \emph{cost} of an answer is the number of relation applications needed to
find the answer. This idea is borrowed from Silvija Seres's work [*].
Now we illustrate the costs of answers by running a miniKanren relation.
Fig.~\ref{def-repeato} defines the relation \texttt{repeato} that 
relates a term \texttt{x} with a list whose elements are all \texttt{x}s.

>>>>

When every sub-goal of a disjunction produces infinite states, the existing 
disjunctive operator allocates half computational effort to its first goal, 
quater to the second, eighth to the third, and so on. The unfairness provides 
both opportunity and burden: miniKanren users can place more frequently used 
goal at the beginning to optimize their programs; however, it might be a 
catastrophe if a goal that generate many useless partial solution is placed 
before more important goals. Seasoned miniKanreners usually know how to utilize 
the unfairness to optimize their programs. However, we believe search strategies 
that is less sensitive to goal order can also be useful to little miniKanreners 
as well as seansoned ones. We propose two such search strategies, balanced 
interleaving DFS (biDFS) and breadth-first search (BFS), and observe how they 
affect the efficiency and the answer order of known miniKanren programs. The 
experiment is conducted with the miniKanren from \textit{The Reasoned Schemer, 
2nd Edition}.

\subsection{fairness}

A disjunction is fair if when a corresponding goal is queried, answers of 
lower costs come first. The \emph{cost} of an answer the number of relation 
applications needed to find the answer. This idea is borrowed from Silvija 
Seres's work [*].

Interleaving DFS, the search strategy of TRS2 miniKanren, is not fair. Among 
the search strategies we devised, biDFS is not fair, but is more fair than 
iDFS. And BFS is fair.

\subsection{related works}

The unfairness of disjunction has been noticed by Seres et al. Their complete 
and fair search strategy is also named breadth-first search. Their BFS 
is fair and is similar to ours. However, their Haskell implementation cannot be 
translated to Scheme directly. This is partially due to the difference if 
calling convensions of host languages. Another reason is that their search 
space are infinite even when no answers exists. This is not feasible 
in miniKanren, where users can query all answers.

% TODO: summarize following sections) Section 2 explains why iDFS biases toward 
% left \texttt{conde} clauses. Section 3 is about the biDFS. Section 4 is 
% about BFS.

\section{representation of search space}

``Stream'' is often used to name the representation of search space in 
miniKanren. However, the search space is more like a stream of list, 
because the information that a stream has no \texttt{car} is employed in all 
complete miniKanren search strategies to redirect search effort. If the search 
space is a normal stream, the information should have been used in a more 
trivial way.

In the rest of this paper, we call the \texttt{car}s of a stream 
its \textit{mature part}, and the last \texttt{cdr} its \textit{immature part}.

\section{why disj is unfair}

The multiarity disjuction operator \texttt{disj} combine its sub-goals with 
\texttt{disj2}, right associatively. \texttt{disj2} is fair in the sence that 
it allocates computational effort evenly to its two sub-goals. Unfortunately, 
the way how \texttt{disj2} achieves fairness leads to the unfairness of 
\texttt{disj}. The core functionality of \texttt{disj2} is completed by 
\texttt{append-inf} (Fig.~ref{disj2-and-append-inf}). When both input streams 
are infinite, the resulting mature part contains only the mature part of 
\texttt{s-inf}. The whole \texttt{t-inf} goes to the resulting immature part. 
However, \texttt{t-inf} and \texttt{s-inf} are swapped in the delayed recursive 
call. Hence the search strategy spend computational effort evenly in two 
appended streams. As \texttt{disj} applys \texttt{disj2} right associatively, 
the left clauses are more frequently explored.

@@

In this section we change the search strategy to \textit{breadth-first search} 
and optimize it. The whole process is completed in two steps. In the first 
step, from mk-0 to mk-1, BFS is introduced. In the second step, mk-1 to mk-3, 
BFS is optimized. The initial version, \texttt{mk-0}, is exactly the version in 
\emph{The Reasoned Schemer, 2nd Edition}.

@@

Now we illustrate the costs of answers by running a miniKanren relation.
Fig.~\ref{def-repeato} defines the relation \texttt{repeato} that 
relates a term \texttt{x} with a list whose elements are all \texttt{x}s.
